---
title: "mlrCPO Core"
author: "Martin Binder"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. mlrCPO Core}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, eval = TRUE, child = 'toc/vignettetoc.Rmd'}
```

## CPO

**CPO**s are first-class objects in R that represent data manipulation. They can be combined to form networks of operation, they can be attached to `mlr` `Learner`s, and they have tunable Hyperparameters that influence their behaviour.

### Preparation

```{r}
library("mlrCPO")
df = data.frame(a = 1:3, b = -(1:3) * 10)
```

## Lifecycle of a CPO

### CPO Constructor

```{r}
print(cpoPca)  # example CPOConstructor
class(cpoPca)
```

CPO constructors have parameters that

* set the CPO Hyperparameters
* set the CPO ID (default NULL)
* resetrict the data columns a CPO operates on (`affect.*` parameters)

```{r}
names(formals(cpoPca))
print(cpoPca, verbose = TRUE)  # verbose print includes function info
```

### CPO

```{r}
(cpo = cpoScale()) # construct CPO with default Hyperparameter values
class(cpo)  # CPOs that are not compound are "CPOPrimitive"
getCPOClass(cpo)  # CPO Class: CPO, CPOInverter, CPORetrafo, NULLCPO
getCPOOperatingType(cpo)  # Operating on feature, target, both?
print(cpo, verbose = TRUE)  # detailed printing
```

#### Functions that work on CPOs:
```{r}
getParamSet(cpo)
getHyperPars(cpo)
setHyperPars(cpo, scale.center = FALSE)
getCPOId(cpo)
setCPOId(cpo, "MYID")
getCPOName(cpo)
getCPOAffect(cpo)  # empty, since no affect set
getCPOAffect(cpoPca(affect.pattern = "Width$"))
getCPOProperties(cpo)  # see properties explanation below
getCPOPredictType(cpo)
getCPOConstructor(cpo)  # the constructor used to create the CPO
```

#### Exporting Parameters

Sometimes when using many CPOs, their hyperparameters may get messy. CPO enables the user to control which hyperparameter get exported. The parameter "export" can be one of "export.default", "export.set", "export.unset", "export.default.set", "export.default.unset", "export.all", "export.none". "all" and "none" do what one expects; "default" exports the "recommended" parameters; "set" and "unset" export the values that have not been set, or only the values that were set (and are not left as default). "default.set" and "default.unset" work as "set" and "unset", but restricted to the default exported parameters.

```{r}
(sc = cpoScale())
getParamSet(sc)
```
```{r}
(sc = cpoScale(export = "export.none"))
getParamSet(sc)
```
```{r}
(sc = cpoScale(scale = FALSE, export = "export.unset"))
getParamSet(sc)
```

#### CPO Application using `%>>%` or `applyCPO`
`CPO`s can be applied to `data.frame` and `Task` objects.

```{r}
head(iris) %>>% cpoPca()
head(getTaskData(applyCPO(cpoPca(), iris.task)))
```

#### CPO Composition using `%>>%` or `composeCPO`
`CPO` composition results in a new CPO which mostly behaves like a primitive CPO. Exceptions are:

* Compound CPOs have no `id`
* Affect of compound CPOs cannot be retrieved

```{r}
scale1 = cpoScale()
scale2 = cpoScale()
```
```{r, error = TRUE}
scale1 %>>% scale2  # error! parameters 'center' and 'scale' occur in both
```
```{r}
compound = setCPOId(scale1, "scale1") %>>% setCPOId(scale2, "scale2")
composeCPO(setCPOId(scale1, "scale1"), setCPOId(scale2, "scale2"))  # same
class(compound)
print(compound, verbose = TRUE)
getCPOName(compound)
getParamSet(compound)
getHyperPars(compound)
setHyperPars(compound, scale1.center = TRUE, scale2.center = FALSE)
```
```{r, error = TRUE}
getCPOId(compound)  # error: no ID for compound CPOs
getCPOAffect(compound)  # error: no affect for compound CPOs
```

#### Compound CPO decomposition, CPO chaining

```{r}
as.list(compound)
pipeCPO(as.list(compound))  # chainCPO: list CPO -> CPO
```

#### CPO - Learner attachment using `%>>%` or `attachCPO`

```{r}
lrn = makeLearner("classif.logreg")
(cpolrn = cpo %>>% lrn)  # the new learner has the CPO hyperparameters
attachCPO(compound, lrn)  # attaching compound CPO
```

```{r}
# CPO learner decomposition
getLearnerCPO(cpolrn)  # the CPO
getLearnerBare(cpolrn)  # the Learner
```

### Retrafo
CPOs perform data-dependent operation. However, when this operation becomes part of a machine-learning process, the operation on predict-data must depend only on the training data.

The `Retrafo` object represents the re-application of a trained CPO

```{r}
transformed = iris %>>% cpo
head(transformed)
(ret = retrafo(transformed))
```
```{r}
# General methods that work on retrafo
getCPOName(ret)
getHyperPars(ret)
getCPOClass(ret)  # CPO Class: CPO, CPOInverter, CPORetrafo, NULLCPO
getCPOOperatingType(ret)  # Operating on feature, target, both?
getCPOPredictType(ret)
getCPOTrainedCapability(ret)  # can this be used for retrafo and/or invert?
```
```{r}
# retrafos are stored as attributes
attributes(transformed)$retrafo
```

#### Retrafo Inspection
`Retrafo` objects can be inspected using `getCPOTrainedState`. The state contains the hyperparameters, the `control` object (CPO dependent data representing the data information needed to re-apply the operation), and information about the `Task` / `data.frame` layout used for training (column names, column types) in `data$shapeinfo.input` and `data$shapeinfo.output`.

The state can be manipulated and used to create new `Retrafo`s, using `makeCPOTrainedFromState`.

```{r}
(state = getCPOTrainedState(retrafo(iris %>>% cpoScale())))
state$control$center[1] = 1000  # will now subtract 1000 from the first column
new.retrafo = makeCPOTrainedFromState(cpoScale, state)
head(iris %>>% new.retrafo)
```

#### Application of Retrafo using `%>>%`, `applyCPO`, or `predict`

```{r}
head(iris) %>>% retrafo(transformed)
```
Should give the same as head(transformed), since the same data was used:
```{r, eval = FALSE}
# same:
applyCPO(retrafo(transformed), head(iris))
predict(retrafo(transformed), head(iris))
```

#### Retrafos are automatically chained when applying CPOs (!!!)
When executing `data %>>% CPO`, the result has an associated `Retrafo` object. When applying another `CPO`, the `Retrafo` will be the chained operation. This is to make `data %>>% CPO1 %>>% CPO2` the way one expects it to work.

```{r}
data = head(iris) %>>% cpoPca()
retrafo(data)
data2 = data %>>% cpoScale()
# retrafo(data2) is the same as retrafo(data %>>% pca %>>% scale)
retrafo(data2)
# to interrupt this chain, set retrafo to NULL
retrafo(data) = NULL
data2 = data %>>% cpoScale()
retrafo(data2)
```

#### Retrafo Composition, Decomposition, Chaining
Using `as.list` and `pipeCPO`, just like for `CPO`s.
```{r}
compound.retrafo = retrafo(head(iris) %>>% compound)
compound.retrafo
(retrafolist = as.list(compound.retrafo))
retrafolist[[1]] %>>% retrafolist[[2]]
pipeCPO(retrafolist)
```

### Inverter
Inverters represent the operation of inverting transformations done to prediction columns. They are not usually exposed outside of `Learner` objects, but can be retrieved when retransformed data is tagged using `tagInverse`.

Inverters are not yet fully implemented.

```{r, eval = FALSE}
# there is currently no example targetbound cpo
logtransform = makeCPOTargetOp("logtransform",
  properties.target = "regr", constant.invert = TRUE,
  cpo.train = NULL,
  cpo.train.invert = NULL,
  cpo.retrafo = {
    target[[1]] = log(target[[1]])
    target
  }, cpo.invert = { exp(target) })


log.retrafo = retrafo(bh.task %>>% logtransform())  # get a target-bound retrafo
getCPOKind(log.retrafo)  # logtransform is *stateless*, so it is a retrafo *and* an inverter
getCPOBound(log.retrafo)

inverter(bh.task %>>% log.retrafo)
```
```{r, eval = FALSE}
inverter(tagInvert(bh.task) %>>% log.retrafo)
```

Inverting is done with the `invert` function.

```{r, eval = FALSE}
log.bh = bh.task %>>% logtransform()
log.prediction = predict(train("regr.lm", log.bh), log.bh)
invert(retrafo(log.bh), log.prediction)  # not implemented :-/
invert(retrafo(log.bh), log.prediction$data["response"])  # not implemented :-/
```

## CPO Properties
CPOs contain information about the kind of data they can work with, and what kind of data they produce. `getCPOProperties` returns a list with the slots `handling`, `adding`, `needed`, `properties$adding`, indicating the kind of data a CPO can handle, the kind of data it needs the data receiver (e.g. attached learner) to have, and the properties it adds to a given learner. An example is a CPO that converts factors to numerics: The receiving learner needs to handle numerics, so `properties$needed = "numerics"`, but it *adds* the ability to handle factors (since they are converted), so `properties$adding = c("factors", "ordered")`. `properties$data` is only different from `properties$handling` if `affect.*` parameters are given. In that case, `properties$data` determines what properties the selected subset of columns must have.

```{r, error = TRUE}
getCPOProperties(cpoDummyEncode())
train("classif.geoDA", bc.task)  # gives an error
```
```{r}
train(cpoDummyEncode(reference.cat = TRUE) %>>% makeLearner("classif.geoDA"), bc.task)
getLearnerProperties("classif.geoDA")
getLearnerProperties(cpoDummyEncode(TRUE) %>>% makeLearner("classif.geoDA"))
```

